Foundations of ecological and evolutionary change

### 2025-06-18 foundations_of_ecoevo_arxiv.Rmd
Edits for resubmission of revision to ecology and evolution

---
title: "Foundations of ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  word_document: default
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
bibliography: refs.bib
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
biblio-style: apalike
---

@@

Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density. 
The deep interdependence between ecology and evolution is well-established, and recent models have focused on integrating eco-evolutionary dynamics. 
Nevertheless, a gap remains between the logical foundations of ecology and evolution. 
We bridge this gap with an equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived. 
We thereby identify an overlooked bridge between the fundamental equations of population ecology and biological evolution. 
Our unification formally recovers the equivalence between mean population growth rate and evolutionary fitness and links this change to ecosystem function. 
We outline how our proposed framework can be used to further develop eco-evolutionary theory.

@@

Theoretical unification is a powerful tool for scientific advancement.
Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions. 
In evolutionary biology, the Price equation (Box 1) provides a unifying framework for evolutionary theory by exhaustively and exactly describing evolutionary change for any closed evolving population [@Price1970; @Luque2016; @Lehtonen2020].
The Price equation is therefore fundamental, in the sense that it binds together all of evolutionary theory by formally defining what evolutionary change is and is not [@Price1970; @Rice2004; @Gardner2008; @Frank2015a; @Luque2016; @Luque2021].
Using this formal definition, the scope of, and relationships among, sub-disciplines within evolutionary theory can be clarified.
For example, fundamental equations of both population and quantitative genetics can be derived from the Price equation [@Queller2017].
This provides conceptual clarity by demonstrating the logical consistency of different theoretical frameworks within evolutionary biology.
Our aim here is to propose an equation that extends this conceptual clarity to include population size change and thereby provide a formal and exact definition for joint ecological and evolutionary change.

@@

We present an equation from which the fundamental equations of ecology and evolutionary biology can be derived.
Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics (e.g., the average excess) or quantitative genetics (e.g., the Breeder's equation) can be derived from the Price equation [@Queller2017].
We propose our equation as a formal definition of eco-evolutionary change.

@@

This foundation of eco-evolutionary change is an interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
The value of $\Omega$ is a summation of individual characteristics and takes the same units as $z$.
From eqn 1, we can derive the most fundamental equations of population ecology and evolutionary biology through an appropriate interpretation of $z$.
Under more limited interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.

@@

To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
In this case,

@@

We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$.
Summing up $\beta_{i}$, $\delta_{i}$, and current entities tallies up the total number of entities in the next time step,

@@

Now we must consider the total conservation of probability [@Frank2015; @Frank2016].
In eqn 7, $\Omega$ is the total sum trait values ($z_{i}$) across the entire population at $t + 1$ divided by the number of individuals ($N$) in the population at $t$.
But the size of the population can change from $t$ to $t + 1$.
To recover mean trait change for the Price equation (and therefore conserve total probability), we need to account for this change in population size.
The left-hand side of eqn 7 describes contributions to the sum trait value from the new population at $t + 1$.
But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weigh $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.

@@

From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).

@@

In some cases, $\Omega$ could also be intepreted as the total contribution of a population to ecosystem function.
This is restricted to cases in which $z$ is a characteristic defining an absolute quantity measured at the whole organism level such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
In such cases, the sum across individuals gives a meaningful total quantity for the population.
When $z$ is instead defined by relative organism-level measurements such wing loading, nutrient ratio, or diet composition, or when $z$ is measured at a level of biological organisation below the organism (e.g., average cell volume or leaf surface area), $\Omega$ does not have a clear population-level interpretation.

Box 3 provides a toy example of three plants with different fitnesses and carbon captures.
We use this toy example to demonstrate a concrete example of our proposed framework.

@@

> **Box 3:** As a toy example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual total fruit mass (kg) is measured, and change is observed over a year. For all plants, $\delta_{i} = 1$, and let plant fecundities be $\beta_{1} = 1$, $\beta_{2} = 1$, and $\beta_{3} = 2$. Applying eqn 1 to population change such that $\Omega = N_{t+1}$, $z_{i} = 1$ and $\Delta z_{i} = 0$, $N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4$ (note $\bar{w} = 4/3$, so $N_{t}\bar{w} = 4$). Focusing next on the characteristic of total fruit mass, let $z_{1} = 0.8$ kg, $z_{2} = 1.0$ kg, and $z_{3} = 1.5$ kg. Also let $\Delta z_{i} = 0.1$ for all plants to reflect a change in soil environment from $t$ to $t + 1$. In this case, population fruit yield at $t+1$ is $\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2$ kg. At $t$, mean fruit yield per plant was 1.1 kg, but at $t+1$, mean fruit yield per plant is $1.3$ kg. Note that $\mathrm{Cov}(w, z) = 2/15$ and $\mathrm{E}(w \Delta z) = 2/15$, so applying the Price equation, $\bar{w}\Delta z = 2/15 + 2/15 = 4/15$. Since $\bar{w} = 4/3$, multiplying both sides of the equation by $3/4$ returns $\Delta z = 0.2$, which is the mean difference in fruit yield between $t+1$ and $t$. The framework expressed in eqn 1 thereby links population change, evolutionary change, and ecosystem function.

@@

An important aspect of scientific progress is the ability to connect disparate theories and models to show how specific empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961; @Morrison2000].
Rather than making simplifying assumptions, as is the approach for specific ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 1).
Our eqn 1 thereby provides a foundation for defining eco-evolutionary change in any population.

The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
Because of its abstract nature and lack of any system-specific assumptions, the Price equation is not dynamically sufficient and makes no predictions about what will happen in any particular system [@Gardner2020].
Its role is not to predict, but to formally and completely define and separate components of evolutionary change [@Baravalle2024]. 
The same is true of the general equation for population change (eqn 2), at least as we have used it here where it serves to define what population growth means in ecology.
This equation formally and completely describes population change in terms of births and deaths.
In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population, but one that is revealed to be necessarily abstract and not dynamically sufficient.
We hope that this will be useful for eco-evolutionary theory in a similar way that the Price equation is useful for evolutionary theory: potentially facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.

@@

>>>>

---
title: "Foundations of ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lineno}
linestretch: 1
bibliography: refs.bib
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
biblio-style: apalike
---

@@

Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density. 
The deep interdependence between ecology and evolution is well-established, and recent models focus on integrating eco-evolutionary dynamics to demonstrate how ecological and evolutionary processes interact and feed back upon each other. 
Nevertheless, a gap remains between the logical foundations of ecology and evolution. 
Both population ecology and evolution have fundamental equations that define how the size of a population (ecology) and the average characteristic within a population (evolution) change, respectively, over time.
These fundamental equations are a complete and exact description of change for any closed population, but how they are formally linked remains unclear.
We link the fundamental equations of population ecology and evolution with an equation that sums how individual characteristics interact with individual fitness in a population.
From this equation, we derive the fundamental equations of population ecology and evolutionary biology (the Price equation). 
We thereby identify an overlooked bridge between ecology and biological evolution. 
Our unification formally recovers the equivalence between mean population growth rate and evolutionary fitness and links this change to ecosystem function. 
We outline how our framework can be used to further develop eco-evolutionary theory.

@@

Theoretical unification is a powerful tool for scientific advancement.
Such unification has been a major goal in scientific research throughout history [@Smocovitis1992; @Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions. 
In evolutionary biology, the Price equation (Box 1) provides a unifying framework for evolutionary theory by exhaustively and exactly describing evolutionary change for any closed evolving population [@Price1970; @Luque2016; @Lehtonen2020].
The Price equation is therefore fundamental, in the sense that it binds together all of evolutionary theory by formally defining what evolutionary change is and is not [@Price1970; @Rice2004; @Gardner2008; @Frank2015a; @Luque2016; @Luque2021].
Using this formal definition, the scope of, and relationships among, sub-disciplines within evolutionary theory can be clarified.
For example, fundamental equations of both population and quantitative genetics can be derived from the Price equation [@Queller2017].
This provides conceptual clarity by demonstrating the logical consistency of different theoretical frameworks within evolutionary biology.
Our aim here is to propose an equation that extends this conceptual clarity to include population size change and thereby provide a formal and exact definition for joint ecological and evolutionary change.

@@

We present an equation from which the fundamental equations of ecology and evolutionary biology can be derived.
Derivation follows by adding assumptions that are specific to population ecology or evolution in the same way that key equations of population genetics or quantitative genetics can be derived from the Price equation by restricting the domain of interest [e.g., to allele frequencies in the case of population genetics, or to continuous phenotypes in the case of quantitative genetics, @Queller2017].
We propose our equation as a formal definition of eco-evolutionary change.

@@

This foundation of eco-evolutionary change is a statistical interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
The value of $\Omega$ is a summation of individual characteristics and takes the same units as $z$.
From eqn 1, we can derive the most fundamental equations of population ecology and evolutionary biology through an appropriate interpretation of $z$.
Under more limited interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.

@@

To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
In other words, we set $z_{i} = 1$ to simply indicate that $i$ is one member of the population.
In this restricted case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit, see @NewellEtAl2019].
Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
In this case,

@@

We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$.
Summing up $\beta_{i}$, $\delta_{i}$, and current individuals tallies up the total number of individuals in the next time step,

@@

Now we must consider the total conservation of probability [@Frank2015; @Frank2016].
In eqn 7, $\Omega$ is the total sum trait values ($z_{i}$) across the entire population at $t + 1$ divided by the number of individuals ($N$) in the population at $t$.
But the size of the population can change from $t$ to $t + 1$.
To recover mean trait change for the Price equation (and therefore conserve total probability), we need to account for this change in population size.
The left-hand side of eqn 7 describes contributions to the sum trait value from the new population at $t + 1$.
But we cannot treat $\Omega/N$ as the mean of $z$ at $t+1$ ($\bar{z}'$) because we need to weight $N$ by the mean fitness of the population at $t$ to account for any change in population size from $t$ to $t+1$.

@@

Equation 10 is the Price equation.
From eqn 1, which describes fundamental birth and death processes in a population, we can therefore derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).

@@

In some cases, $\Omega$ could also be interpreted as the total contribution of a population to ecosystem function.
This is restricted to cases in which $z$ is a characteristic defining an absolute quantity measured at the whole organism level such as biomass, seed production, carbon capture, flower visits, or nutrient consumption.
In such cases, the sum across individuals gives a meaningful total quantity for the population (e.g., the total biomass or seeds produced in the population).
When $z$ is instead defined by relative organism-level measurements such wing loading, nutrient ratio, or diet composition, or when $z$ is measured at a level of biological organisation below the organism (e.g., average cell volume or leaf surface area), $\Omega$ does not have a clear population-level interpretation.
In such cases, eqn 1 still defines eco-evolutionary change; the interpretation of $\Omega$ by itself is just not as interesting, biologically.

Box 3 provides an instructive example of three plants with different fitnesses and carbon captures.
We use this example to demonstrate a concrete example of our proposed framework.

@@

> **Box 3:** As an instructive example of our framework, consider a population of $N_{t} = 3$ annual plants in which individual total fruit mass (kg) is measured, and change is observed over a year. For all plants, $\delta_{i} = 1$ (because annual plants die between $t$ and $t+1$), and let plant fecundities be $\beta_{1} = 1$, $\beta_{2} = 1$, and $\beta_{3} = 2$. Applying eqn 1 to population change such that $\Omega = N_{t+1}$, $z_{i} = 1$ and $\Delta z_{i} = 0$, $N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4$ (note $\bar{w} = 4/3$, so $N_{t}\bar{w} = 3(4/3) = 4$). Focusing next on the characteristic of total fruit mass, let $z_{1} = 0.8$ kg, $z_{2} = 1.0$ kg, and $z_{3} = 1.5$ kg. Also let $\Delta z_{i} = 0.1$ for all plants to reflect a change in soil environment from $t$ to $t + 1$. In this case, population fruit yield at $t+1$ is $\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2$ kg. At $t$, mean fruit yield per plant was $(0.8 + 1.0 + 1.5)/3 = 1.1$ kg, but at $t+1$, mean fruit yield per plant is $[(0.8 + 0.1) + (1.0 + 0.1) + 2(1.5 + 0.1)]/4 = 1.3$ kg. Note that $\mathrm{Cov}(w, z) = (1/N)\sum_{i}^{N}(z_{i} - \bar{z})(w_{i} - \bar{w}) = (1/3)\left[(0.8 - 1.1)(1-4/3) + (1 - 1.1)(1-4/3) + (1.5 - 1.1)(2-4/3) \right]  = 2/15$ and $\mathrm{E}(w \Delta z) = 4/3 \times 0.1 = 2/15$, so applying the Price equation, $\bar{w}\Delta \bar{z} = 2/15 + 2/15 = 4/15$. Since $\bar{w} = 4/3$, multiplying both sides of the equation by $3/4$ returns $\Delta z = 0.2$, which is the mean difference in fruit yield between $t+1$ and $t$. The framework expressed in eqn 1 thereby links population change, evolutionary change, and ecosystem function.

@@

An important aspect of scientific progress is the ability to connect disparate theories and models to show how specific empirical and theoretical models are logical (mathematical) consequences of more fundamental ones [@Nagel1961; @Morrison2000].
Rather than making simplifying assumptions, as is the approach for specific ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
From the basic axiom that each individual is one member of a specific population ($z_{i} = 1$ and $\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 2) and the principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution, the Price equation (Box 1).
Our eqn 1 thereby provides a foundation for defining eco-evolutionary change in any population.

The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
Because of its abstract nature and lack of any system-specific assumptions, the Price equation is not dynamically sufficient and makes no predictions about what will happen in any particular system [@Gardner2020].
Its role is not to predict, but to formally and completely define and separate components of evolutionary change [@Baravalle2024]. 
The same is true of the general equation for population change (eqn 2), at least as we have used it here where it serves to define what population growth means in ecology.
This equation formally and completely describes population change in terms of births and deaths.
In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population.
Like all fundamental equations, our equation is necessarily abstract and not dynamically sufficient.
We believe that it will be useful for eco-evolutionary theory in a similar way that the Price equation is useful for evolutionary theory: potentially facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements (see below and Supporting Information S1).

@@

<div id="refs"></div>

\clearpage

Supporting Information S1
================================================

This supporting information demonstrates how to derive well-established models in population ecology and evolutionary biology from equation 1 in the main text,

In the main text, we derived both the Price equation and the birth-death model from the above.
Here we integrate interactions between individuals to recover density-dependent population growth, and we integrate groups within the population to recover multi-level selection.
Finally, we integrate both to model a system in which multi-level selection and density dependent population change occur simultaneously.
We can do this in a way that clarifies model assumptions by allowing an individual $j$ to modulate the birth or death of the focal individual $i$.

Density-dependent population growth
===========================================

There are two potential ways to model the incorporation of density dependence into population growth.
We start with what is likely the most familiar model focusing on individual growth rate $r_{i}$, then use a slightly different model focusing on fitness $w_{i}$. 
First note that here we set $\Omega = N_{t+1}$, and $z_{i} = 1$ and $\Delta z_{i} = 0$ for all individuals as in the main text.
We can define $r_{i} = \beta_{i} - \delta_{i}$ as an individual growth rate for $i$ [@Lion2018; @Lion2023].
In this case,

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i} + 1\right)
\tag{S1}
$$

Mathematically, the most general approach here would be to define individual growth as a function of the entire system $\mathbf{E}$, $r_{i}(\mathbf{E})$, where $\mathbf{E}$ is a vector with elements including any parameters potentially relevant to $r_{i}$.
Taking this approach would recover a version of eqn 2 in @Lion2018 and permit any relationship between the system and a focal individual's growth.
Limiting our focus to the effects of other individuals ($j$) and assuming that the effects of these individuals are additive, let $a_{ij|\cdot}$ be the effect of individual $j$ on the growth rate attributable to $i$ conditioned on all other individuals within the population such that $r_{i}\left(1 + \sum_{i = j}^{N}a_{ij|\cdot} \right)$ defines the realised growth rate of $i$,

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij|\cdot} \right) + 1\right).
\tag{S2}
$$

Assuming that individual effects of $j$ on $i$ are also independent, we can remove the condition,

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - \sum_{i = j}^{N_{t}}a_{ij} \right) + 1\right).$$

Further assuming that all individuals have the same per capita effect such that $a = a_{ij}$ for any $i$ and $j$ pair (as might be reasonable given resource competition in a well-mixed population),

$$N_{t+1} = \sum_{i=1}^{N_{t}}\left(r_{i}\left(1 - a N_{t} \right) + 1\right).$$

If $r_{i}$ values are identical,

$$N_{t+1} = N_{t} + r N_{t}\left(1 - a N_{t} \right).
\tag{S3}
$$

Equation S3 therefore recovers a classic version of a discrete time logistic growth by making assumptions from an exact model of eco-evolutionary change.

An alternative approach would be to define model the effects of an individual $j$ on the fitness of $i$ ($w_{i}$), thereby replacing eqn S1 with $N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}$ and replacing eqn S2 with,

$$N_{t+1} = \sum_{i=1}^{N_{t}}w_{i}\left(1 - \sum_{i = j}^{N_{t}}\alpha_{ij|\cdot} \right).
$$

Note that we have used $\alpha_{ij}$ to represent the effect of $j$ on the fitness of $i$ for clarity in the sections below.
By making the same assumptions of additivity, independence, and identical effects such that $\alpha = \alpha_{ij}$ for all $j$ on $i$, and assuming fitness is equal ($w_{i} = \bar{w}$), we can derive,

$$N_{t+1} = \bar{w}N_{t}(1 - \alpha N_{t}).
\tag{S4}
$$

This is an alternative way to express logistic growth.

Multi-level selection
=============================

We can recover multi-level selection from our eqn 1. 
Here we derive the original form of the multi-level @Price1972 equation as it appears in eqn 3.1 of @Lehtonen2020a.
Individuals belong to one of $K$ total groups where $j$ indexes $K$ groups and $i$ indexes individuals. 
Individuals do not overlap in group membership. 
The size of group $j$ is denoted as $N_{j}$.
Equation S5 below uses summations to partition how individuals within each group contribute to $\Omega$,

$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}\left(\beta_{j,i} - \delta_{j,i} + 1 \right)\left(z_{j, i} + \Delta z_{j, i}\right).
\tag{S5}
$$

In S5, indices $\beta_{j, i}$, $\delta_{j, i}$, and $z_{j, i}$ identify individual $i$ in group $j$.
We set $w_{j, i} = \beta_{j, i} - \delta_{j, i} + 1$, and for simplicity let $\Delta z_{j, i} = 0$ (i.e., no transmission bias),

$$\Omega = \sum_{j=1}^{K}\sum_{i=1}^{N_{j}}w_{j,i}z_{j, i}.$$

For ease of presentation, with no loss of generality, we assume all group sizes are equal with a group size of $N_{j} = n$ for all $j$.
If group sizes differ, then weighted expectations and covariances are instead needed [@Lehtonen2020a].
Given equal group sizes, the total number of individuals ($N$) equals $K \times n$, and,

$$\frac{\Omega}{K n} = \left(\frac{1}{K}\right)\left(\frac{1}{n}\right)\sum_{j=1}^{K}\sum_{i=1}^{n}w_{j,i}z_{j, i}.$$

Rearranging,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\frac{1}{n}\sum_{i=1}^{n}w_{j,i}z_{j, i}.$$

The inner summation can be rewritten as an expectation for group $j$,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i} z_{j,i}\right).$$

As in the main text, we note $\mathrm{E}(XY) = \mathrm{Cov}(X,Y) + \mathrm{E}(X)\mathrm{E}(Y)$.
Defining $Cov_{j}(w_{j}, z_{j})$ as the covariance between $w_{j,i}$ and $z_{j, i}$ for group $j$,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).$$

We can separate the summation for each term,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \frac{1}{K}\sum_{j=1}^{K}\mathrm{E}_{j}\left(w_{j,i}\right)\mathrm{E}_{j}\left(z_{j,i}\right).$$

Using the notation $\bar{w}_{j} = E_{j}(w_{j,i})$ and $\bar{z}_{j} = E_{j}(z_{j,i})$ to indicate the expectation in group $j$,

$$\frac{\Omega}{K n} = \frac{1}{K}\sum_{j=1}^{K}\mathrm{Cov}_{j}\left(w_{j,i}, z_{j,i}\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).$$

We can also rewrite the first term on the right-hand side as an expectation,

$$\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) + \mathrm{E}\left(\bar{w}_{j} \bar{z}_{j} \right).$$

We can rearrange the second term on the right-hand side ($\bar{\bar{w}}$ indicates grand mean over all groups),

$$\frac{\Omega}{K n} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \bar{\bar{w}}\bar{\bar{z}}.$$

As in the main text, note that $Kn\bar{\bar{w}}$ accounts for differences in total population size from $t$ to $t+1$, with $\bar{\bar{w}}$ being mean fitness across all groups.
We can therefore set $\Omega = Kn\bar{\bar{w}}\bar{\bar{z'}}$, so,

$$\frac{Kn\bar{\bar{w}}\bar{\bar{z'}}}{K n} - \bar{\bar{w}}\bar{\bar{z}} = \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right) +  \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right).$$

Because $\Delta \bar{\bar{z}} = \bar{\bar{z'}} - \bar{\bar{z}}$,

$$\bar{\bar{w}}\Delta \bar{\bar{z}} = \mathrm{Cov}\left(\bar{w}_{j}, \bar{z}_{j} \right) + \mathrm{E}\left(\mathrm{Cov}_{j}\left(w_{j}, z_{j}\right)\right).
\tag{S6}
$$

This recovers the multi-level @Price1972 equation [@Lehtonen2020a] from a starting point of eco-evolutionary change in different groups.
Equation S6 can be found in @Lehtonen2016 B2.I, who then derives a multi-level selection version of Hamilton's rule predicting the evolution of altruism.

Integration of ecology and evolution
========================================

For simplicity, we now focus on showing an integration between ecology and evolution using a population with no multi-level selection and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias).
As above in the section on density-dependent population growth, we define $w_{i} = \beta_{i} - \delta_{i} + 1$ and use $\alpha_{i,j}$ to represent the effect of $j$ on the fitness of $i$.
Our starting equation is therefore,

$$\Omega = \sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
\tag{S7}
$$

We have already demonstrated that if we assume all individuals have the same effect on a focal individual such that $\alpha = \alpha_{i,j}$ for all $i$ and $j$ pairs, we can recover equation S4 when $z_{i} = 1$ and $\Omega$ is therefore interpreted as the count of entities,

$$N_{t+1} = \bar{w} N_{t} \left(1 - \alpha N_{t}\right).$$

We now start from S7 to derive $\Delta \bar{z}$.
The objective is to use our definition of eco-evolutionary change to simultaneously recover how interactions between individuals affect population change and evolutionary change.

We start by dividing both sides of S7 by $N$,

$$\frac{\Omega}{N} = \frac{1}{N}\sum_{i=1}^{N}w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}.
\tag{S8}
$$

We can express the right-hand side of eqn S8 as an expectation,

$$\frac{\Omega}{N} = \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)z_{i}\right).
\tag{S9}
$$

We can rewrite the right-hand side in terms of covariances,

$$\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \mathrm{E}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)\right)\mathrm{E}\left(z_{i}\right). 
\tag{S9}
$$

The expectations in the second term on the right-hand side can be replaced with overbars to represent the mean,

$$\frac{\Omega}{N} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right) + \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i}. 
\tag{S9}
$$

In the main text, we noted that $\Omega = N\bar{w}\bar{z}'$.
Here mean fitness incorporates individual interactions, therefore,

$$\Omega = N\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}'.
\tag{S10}
$$

We can therefore rewrite S9,

$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\bar{z}' -  \overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}    \bar{z}_{i} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
\tag{S11}
$$

Noting again $\Delta \bar{z} = \bar{z}' - \bar{z}$,

$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right), z_{i}  \right). 
\tag{S12}
$$

We can rewrite the left-hand side of S12,

$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i} - w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
\tag{S13}
$$

The covariance term can be split without any additional assumptions,

$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(w_{i}\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right). 
\tag{S14}
$$

If we are able to further assume that $w_{i}$ and the summation over $\alpha_{i,j}$ are independent, then we could rewrite,

$$\overline{w_{i}\left(1 - \sum_{j=1}^{N}\alpha_{i, j}\right)}\Delta\bar{z} = \mathrm{Cov}\left(w_{i}, z_{i}  \right) - \mathrm{Cov}\left(\sum_{j=1}^{N}\alpha_{i, j}, z_{i}  \right)\bar{w_{i}}. 
\tag{S15}
$$

Partitioning fitness into different components with the Price equation is commonplace.
But this derivation highlights, e.g., the ecological and evolutionary relationship between nonsocial and social components of fitness, and population size.
For example, the second term on the right-hand side of S13 shows the covariance between the sum of social interactions and a trait.
When traits covary with the interaction between sociality and fitness, they will have a stronger effect on trait change.
The magnitude of this second term will also increase with $N$, which reflects the stronger effect of the interaction between sociality and fitness when there are more individuals interacting with a focal individual.

We can start again with $K$ groups and let $\Delta z_{j, i} = 0$ (i.e., no transmission bias).
But for simplicity, now consider the case in which $\delta_{j,i} = 1$ (complete mortality), and consider $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K}\sum_{q=1}^{n} \alpha^{j, i}_{p,q})$, where $\alpha^{j, i}_{p,q}$ is the effect that individual $q$ from group $p$ has on a focal individual $i$ from group $j$.
To keep the model tractable, we assume that all individuals within a group $p$ have identical effects on any focal individual in the group $j$, such that $w_{j, i} = \beta_{j, i}(1 - \sum_{p=1}^{K} \alpha^{j}_{p} n)$
